---
title: "Week 3 Project"
author: ""
date: "2023-02-10"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
```

## NYPD Shooting Data
This project will download the New York Police Department's shooting data from data.gov

```{r clean/transform data}
# Read NYPD Shooting Data from Data.Gov
#data_file = "NYPD_Shooting_Incident_Data__Historic_.csv"

url = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

NYPD_shooting_data = read.csv(url)
summary(NYPD_shooting_data)

# Clean datafrome by converting date strings to datetime and convert time from 
# character to time format
NYPD = NYPD_shooting_data %>% 
       mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>% 
       mutate(OCCUR_TIME = parse_time(OCCUR_TIME))

# Convert booleans to integers
NYPD$STATISTICAL_MURDER_FLAG = as.integer(as.logical(NYPD$STATISTICAL_MURDER_FLAG))

```


```{r organize data}
# Group data using group_by locations, precinct, and jurisdiction
NYPD = NYPD %>% 
  group_by(BORO, PRECINCT, JURISDICTION_CODE) %>% 
  select(OCCUR_DATE, BORO, PRECINCT, JURISDICTION_CODE,
         STATISTICAL_MURDER_FLAG, VIC_SEX, 
         VIC_AGE_GROUP, PERP_SEX, PERP_AGE_GROUP,
         VIC_RACE, PERP_RACE, Longitude, Latitude) %>% 
  ungroup()
```
<br>
## Data Plots
<br>
The first plot illustrates the total number of murders by borough.  The second plot 
illustrates the shooting murder frequency by borough by date.  As one can see, the bar chart 
indicates that the borough of Brooklyn has the highest shooting homicide rate, but this data
is only looking at totals from the years 2005 to 2020.  When looking at the shooting homicides
across dates, one can see how Brooklyn and Bronx boroughs have a high number of 
shooting homicides but that these homicides fluctuate over the years.


```{r data plots, echo=FALSE}
# Plot of shooting homicide by borough
NYPD[NYPD$STATISTICAL_MURDER_FLAG>0,] %>% 
  ggplot(aes(x=STATISTICAL_MURDER_FLAG, y=after_stat(count),col=BORO)) +
  geom_bar(position='dodge2') + 
  ggtitle("Shooting Homicide by borough")

# Plot of shooting homicide frequency
NYPD[NYPD$STATISTICAL_MURDER_FLAG>0,] %>% 
  ggplot(aes(x=OCCUR_DATE, y=after_stat(count),col=BORO)) +
  geom_freqpoly(binwidth=100) + 
  ggtitle(" Shooting Homicide by Date Segregated by borough")

# Plot of shooting homicide by sex
NYPD[NYPD$STATISTICAL_MURDER_FLAG>0,] %>% 
  ggplot(aes(x=VIC_SEX, y=after_stat(count),col=BORO)) +
  geom_bar(position='dodge2') + 
  ggtitle("Shooting Homicide by Sex and borough")

# Plot of shooting homicide by race
NYPD[NYPD$STATISTICAL_MURDER_FLAG>0,] %>% 
  ggplot(aes(x=VIC_RACE, y=after_stat(count),col=BORO)) +
  geom_bar(position='dodge2') + 
  ggtitle("Shooting Homicide by Victim Race and borough") + 
  theme(axis.text.x=element_text(angle=45,hjust=1)) 

# Plot of shooting homicide by location
NYPD[NYPD$STATISTICAL_MURDER_FLAG>0,] %>%
  ggplot(aes(x=Longitude, y=Latitude, col=BORO)) +
  geom_point() + 
  ggtitle("Shooting Homicide by Physical Location")

```
<br>
## Analysis
<br>
In reviewing the previous shown plots, it was observed that there is a large difference
in the number of male shooting homicides versus female shooting homicides.  However,
looking at the shooting homicide by sex and borough, it appears that the female homicides
may correlate to the male homicide.  The following analysis calculates the proportion (ratio of
female to male) of female shooting homicides.  As can be seen in the dot plot below,
the proportion for 3 of the 5 boroughs is approximately 10%.  However, Stanten Island and
Queens have a higher proportion of female shooting homicides to male than Brooklyn, Bronx, and
Manhattan.
``` {r analysis}

BOROS = NYPD %>% distinct(BORO)
#convert tibble to string list
BOROS = BOROS$BORO
M_F_Homicide_proportions = rep(0, length(BOROS))

for(i in 1 : length(BOROS)){
  m = NYPD %>% filter(BORO == BOROS[i] & VIC_SEX == "M") %>% 
    summarise(sum(STATISTICAL_MURDER_FLAG))
  f = NYPD %>% filter(BORO == BOROS[i] & VIC_SEX == "F") %>% 
    summarise(sum(STATISTICAL_MURDER_FLAG))
  #print(f/m)
  M_F_Homicide_proportions[i] = as.numeric(f/m)
}
# female shooting homicide proportions data frame 
f_homicide_df = data.frame(female_homicide_proportions = M_F_Homicide_proportions, BOROS)

ggplot(f_homicide_df, aes(x=c(1:5),y=female_homicide_proportions,col=BOROS)) + 
  geom_point() + 
  ylim(0,0.25) + 
  xlab("borough") + 
  ggtitle("Female to Male Shooting Homicide Proportions")

```
<br>
## Conclusion
<br>
Overall, it appears that Brooklyn and Brox both have the highest total shooting 
homicides.  When looking at the plot of the homicide location (latitude, longitude), 
it appears that Brooklyn may cover a large, densely populated area. Additionally, Bronx
appears to be a densely populated area.  However, additional research into this may be
required to identify the population of those areas and overlay a map of the city to
better identify the boundaries of the boroughs.

In regards to bias, my analysis was driven by looking at location, race, and sex of
the perpetrators and the victims.  Looking through this frame may bias the results as
it doesn't fully consider circumstances, time, and other factors that may affect the
interpretation of the results.

The preliminary research here does not paint a complete picture.  Future work
on this should include a map clearly identifying the boundaries of the boroughs, looking 
at the time of year the homicides occur, and add additional information such as population
of the boroughs and include financial information for the boroughs.
